---
output:
  html_document: default
  pdf_document: default
---

Red Wine Exploration by Oliver A. Weigand
========================================================

This report explores a tidy dataset that contains almost 1,600 red wines with 
11 variables on the chemical properties of the wine.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Packages
library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(rstudioapi)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Note: Rmd working & knit directory are by defualt set to document directory

# read data
wine <- read.csv('wineQualityReds.csv')

```

```{r echo=FALSE, message=FALSE, warning=FALSE, cleaning_dim}

# remove X from wine
wine$X <- NULL

# create pretty wine
pretty.wine <- wine

# create quality factorial
wine$quality.factor <- factor(wine$quality)

```
Some data wrangling has been completed to remove `X` from dataset and to create 
`quality.factor` in the red wine dataset.


# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_detials}

# Dataset general details

dim(wine)
str(wine)
summary(pretty.wine)

# remove pretty.wine
rm(pretty.wine)

```

Our data set consist of 13 variables, with 1,599 observations.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# quality
p1 <- ggplot(data = wine,aes(x = quality)) + 
  geom_bar() +
  scale_x_continuous(breaks = seq(1,10,1))

p2 <- ggplot(data = wine, aes(x = quality)) +
  geom_density(adjust = 3)

grid.arrange(p1,p2, ncol=2)

str(wine$quality)
summary(wine$quality)
```

Quality was determined by a minimum of 3 wine experts, who rated each wine 
between 0 (very bad) and 10 (excellent).

Quality has a normal distribution. Note that the values are integers and that 
minimum and maximum values are 3 and 8 respectively.

A Density Plot is provided on the right to better visualizes the distribution 
of the data. 
A Density Plot is a Histogram that uses kernel smoothing to plot values, 
allowing for smoother distributions by smoothing out the noise. 
The peaks of help display where values are concentrated over the interval.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

## acid block

# Fixed acidity plot
p1 <- ggplot(data = wine, aes(x = fixed.acidity)) + 
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = fixed.acidity)) +
  geom_density()

grid.arrange(p1,p2, ncol=2)

# Volatile acidity plot
p1 <- ggplot(data = wine, aes(x = volatile.acidity)) +
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = volatile.acidity)) +
  geom_density()

grid.arrange(p1,p2, ncol=2)

# Citric acid
p1 <- ggplot(data = wine, aes(x = citric.acid)) +
  geom_histogram(bins = 50)

p2 <- ggplot(data = wine, aes(x = citric.acid)) +
  geom_density()

grid.arrange(p1,p2, ncol=2)

# summary
acid.wine <- wine[,c("fixed.acidity", "volatile.acidity", "citric.acid")]
summary(acid.wine)
rm(acid.wine)

```

Fixed acidity has a normal distribution, while volatile acidity has a bimodal 
distribution. 
Citric acid appears to be multimodal with 3 peaks and a tail to the right. 
All plots have outliers.

Most acids in wine, including fixed or nonvolatile acids are quantified as 
fixed acidity. 
The physicochemical test was for tartaric acid.
Volatile acidity is the amount of acetic acid in wine, which can lead to an 
unpleasant, vinegar taste at high of levels.
Citric acid, found in small quantities, can add 'freshness' and flavor to wines. 
All measurements are in g / dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# Free sulfur dioxide plot
p1 <- ggplot(data = wine, aes(x = free.sulfur.dioxide)) + 
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = free.sulfur.dioxide)) + 
  geom_density()

grid.arrange(p1,p2, ncol=2)

# Total sulfur dioxide plots
p1 <- ggplot(data = wine, aes(x = total.sulfur.dioxide)) + 
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = total.sulfur.dioxide)) + 
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# Sulphates plot
p1 <- ggplot(data = wine,aes(x = sulphates)) + 
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine,aes(x = sulphates)) +
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# summary
sulfar.wine <- wine[, c("free.sulfur.dioxide", 
                        "total.sulfur.dioxide", "sulphates")]
summary(sulfar.wine)
rm(sulfar.wine)

```

Free sulfur dioxide, total sulfur dioxide, and sulphates all have a normal 
distribution with a tail to the right and outliers.

Free sulfur dioxide measures the free form of SO2 that exists in equilibrium 
between molecular SO2 (as a dissolved gas) and bisulfite ion. 
It prevents microbial growth and the oxidation of wine.
Total sulfur dioxide measures the total amount of free and bound forms of SO2. 
In low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine.
Sulphates is a wine additive which can contribute to S02 levels, this acts as 
an antimicrobial and antioxidant. All measurements are in g / dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# Residual sugar plot
p1 <- ggplot(data = wine, aes(x = residual.sugar)) + 
  geom_histogram(bins = 50)

p2 <- ggplot(data = wine, aes(x = residual.sugar)) + 
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# Chlorides plot
p1 <- ggplot(data = wine, aes(x = chlorides)) +
  geom_histogram(bins = 50)

p2 <- ggplot(data = wine, aes(x = chlorides)) +
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# summary
sugar.wine <- wine[, c("residual.sugar", "chlorides")]
summary(sugar.wine)
rm(sugar.wine)

```

Both residual sugar and chlorides have normal distributions with outliers and 
tails to the right.

Residual sugar is the amount of sugar remaining after fermentation stops, it's 
rare to find wines with less than 1 gram / liter and wines with greater than 
45 grams / liter are considered sweet.
Chlorides measure the amount of salt (sodium chloride) in the wine. 
All measurements are in g / dm^3.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# Density plot
p1 <- ggplot(data = wine,aes(x = density)) +
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine,aes(x = density)) +
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# pH plot
p1 <- ggplot(data = wine, aes(x = pH)) + 
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = pH)) + 
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# summary
ph.wine <- wine[, c("density", "pH")]
summary(ph.wine)
rm(ph.wine)

```

Both density and pH are normal distributions with a few outliers.

Density measure the density of the wine. All observation should be very close 
to 1 g / cm^3, the density of water. 
Variations are present due to differences in the percent alcohol and sugar 
content.
pH describes how acidic or basic a wine is on a scale from 0 (very acidic) to 
14 (very basic); most wines are between 3 - 4 on the pH scale.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# Alcohol plot
p1 <- ggplot(data = wine, aes(x = alcohol)) +
  geom_histogram(bins = 40)

p2 <- ggplot(data = wine, aes(x = alcohol)) +
  geom_density()

grid.arrange(p1,p2, ncol = 2)

# summary
summary(wine$alcohol)

```

Alcohol has a normal distribution with a tail to the right.

Alcohol measures the percent alcohol content of the wine, in % by volume.

# Univariate Analysis

### What is the structure of your dataset?

There are 1,599 red wines in the dataset with 11 input variables based on 
physiochemical tests (fixed acidity, volatile acidity, citric acid, residual 
sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, 
sulphates, alcohol), and 1 output variable based on sensory data (quality). 
All variables are numeric expect quality which is an integer. 

Other Observations:

- Most wines have quality rating between 5 and 6.
- Most wines are not considered sweet (they have less than 4 grams / liter of 
residual sugar).
- Most wines fall between 3.2 and 3.4 on the pH scale.
- Most wines have an alcohol content between 9.5 and 11.1%.

### What is / are the main feature(s) of interest in your dataset?

The main feature in the dataset is quality. I’d like to determine which 
features are best for predicting the quality of red wines. I suspect that some 
combination of the input variables can be used to build a predictive model for 
quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Citric acid, residual sugar, chlorides, total sulfur dioxide, and alcohol will 
likely contribute to the quality of red wine.

### Did you create any new variables from existing variables in the dataset?

Yes, I created a new variable called `quality.factor`, which is a ordered factor 
with 6 levels (`”3”, “4”, “5”, “6”, “7”, “8”`) original from the `quality` 
variable.
I decided to keep both `quality` and `quality.factor` because it will give me 
greater flexibility and save time from having to convert from an ordered factor 
to and integer and back throughout the project.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

There was one unusual distribution, citric acid, which was a multimodal 
distribution with 3 peaks. 
The dataset was already tidy, so their was little need for any data wrangling.
However I did remove the variable `X` as it was not needed for this exploration 
and created `quality.factor` for convience.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 12, fig.width= 12}

##===============##
##    ggparis    ##
##===============##
# ggparis plot of all variables
ggpairs(wine[,1:12])

```

Matrix of plots with wine dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_corr_plot}

##==================##
## Correlation Plot ##
##==================##

# Calculate correlation values
data <- cor(wine[,1:11], wine[,12], method = 'pearson')

# convert to dataframe
data <- as.data.frame((data))

# convert rownames to column
data <- tibble::rownames_to_column(data, 'name')

# plot
ggplot(data = data, aes(x = reorder(name, abs(V1)), y = V1, fill = abs(V1))) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  geom_label(aes(label = round(data$V1, 3))) +
  scale_fill_gradient(low = 'red', high = 'green') +
  guides(fill=FALSE) + # removes legend
  labs(x = 'Input Variables', y = 'Pearson Correlation',
       title = 'Correlation between Quality and Input Variables',
       subtitle = 'Input variables are based on physicochemical test 
       from Red Wine dataset.') +
  geom_vline(aes(xintercept = 1.5), colour= '#999999', 
             linetype = 2) +  # horizontal line
  geom_text(aes(x = 1.25, y = 0.2525, label = 'Statistically Insignificant'), 
            colour = '#999999')  # line text

```

Plot of correlation values between quality and input variables.


### Plots with main output variable.

These are plots that will look consider a variable against quality.

**Fixed acidity** 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##===============##
## fixed.acidity ##
##===============##

# histogram
p1 <- ggplot(data = wine, aes(x = fixed.acidity)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = fixed.acidity)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5)

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = fixed.acidity))

```

On the left is a histogram of fixed acidity faceted by quality, while on the 
right is a boxplot combined with a scatter-jitter plot of quality and fixed 
acidity.
After the plots is the Pearson’s product-moment correlation calculation.
You can find the correlation on the very bottom of the calculation output under 
sample estimates: cor 0.1240516
Note that statistically significant level is set to 0.05 or 5%.

Fixed acidity has a weak positive correlation with quality.

**Volatile acidity**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
#    volatile.acidity   ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = volatile.acidity)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = volatile.acidity)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5)

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = volatile.acidity))

```

Volatile acidity has a moderate negative correlation with quality.

**Citric acid**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##      citric.acid     ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = citric.acid)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = citric.acid)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5)

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = citric.acid))

```

Citric acid has a weak positive correlation with quality.

**Free sulfur dioxide** 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##  free.sulfur.dioxide ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = free.sulfur.dioxide)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = free.sulfur.dioxide)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm')

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = free.sulfur.dioxide))

```

Free sulfur dioxide has a weak negative correlation with quality. 

**Total sulfur dioxide**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
## total.sulfur.dioxide ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = total.sulfur.dioxide)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = total.sulfur.dioxide)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm')

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = total.sulfur.dioxide))

```

Total sulfur dioxide has a weak negative correlation with quality. 

**Sulphates**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##       sulphates      ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = sulphates)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = sulphates)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5)

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = sulphates))

```

Sulphates has a weak positive correlation with quality. 

**Residual sugar**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##    residual.sugar    ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = residual.sugar)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = residual.sugar)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/4) +
  scale_x_continuous(breaks = seq(3,8,1))

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = residual.sugar))

```

Residual sugar does not have a statistically significant correlation with 
quality. 

**Chlorides**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##       chlorides      ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = chlorides)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = chlorides)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/4) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm')

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = chlorides))

```

Chlorides has a weak negative correlation with quality. 

**Density** 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##        density       ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = density)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = density)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm')

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = density))

```

Density has a weak negative correlation with quality. 

**pH**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##           pH         ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = pH)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = pH)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/3) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm')

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = pH))

```

pH has a weak negative correlation with quality. 

**Alcohol**

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

##======================##
##       alcohol        ##
##======================##

# histogram
p1 <- ggplot(data = wine, aes(x = alcohol)) +
  geom_histogram(bins = 20) +
  facet_wrap(~wine$quality.factor)

# jitter
p2 <- ggplot(data = wine, aes(x = quality, y = alcohol)) +
  geom_boxplot(aes(group=wine$quality.factor)) +
  geom_jitter(alpha = 1/4) +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5)

grid.arrange(p1,p2, ncol = 2)

# cor.test
with(wine, cor.test(x = quality, y = alcohol))

```

Alcohol has a moderate positive correlation with quality.
 
### Plots without main output variable.

**Volatile acidity and citric acid**

```{r echo=FALSE, message=FALSE, warning=FALSE}

##==================================##
## volatile.acidity and citric.acid ##
##==================================##
# plot
ggplot(data = wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_jitter(alpha = 1/3) +
  geom_smooth(method = 'lm') +
  coord_cartesian(ylim = c(0,1))

# cor.test
with(wine, cor.test(x = volatile.acidity, y = citric.acid))

```

Volatile acidity and citric acid have a moderate negative correlation.

**Volatile acidity and sulphates**

```{r echo=FALSE, message=FALSE, warning=FALSE}

##==================================##
##  volatile.acidity and sulphates  ##
##==================================##
# plot
ggplot(data = wine, aes(x = volatile.acidity, y = sulphates)) +
  geom_jitter(alpha = 1/3) +
  geom_smooth(method = 'lm')

# cor.test
with(wine, cor.test(x = volatile.acidity, y = sulphates))

```

Volatile acidity and sulphates have a weak negative correlation. 

**Volatile acidity and alcohol**

```{r echo=FALSE, message=FALSE, warning=FALSE}

##==================================##
##   volatile.acidity and alcohol   ##
##==================================##
# plot
ggplot(data = wine, aes(x = volatile.acidity, y = alcohol)) +
  geom_jitter(alpha = 1/3) +
  geom_smooth(method = 'lm')

# cor.test
with(wine, cor.test(x = volatile.acidity, y = alcohol))

```

Volatile acidity and alcohol have a weak negative correlation. 

**Citric acid and sulphates**

```{r echo=FALSE, message=FALSE, warning=FALSE}

##==================================##
##     citric.acid and sulphates    ##
##==================================##
# plot
ggplot(data = wine, aes(x = citric.acid, y = sulphates)) +
  geom_jitter(alpha = 1/3) + 
  geom_smooth(method='lm')

# cor.test
with(wine, cor.test(x = citric.acid, y = sulphates))

```

Citric acid and sulphates have a moderate positive correlation.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Residual sugar was the only input not to have a significant correlation with the 
main feature (quality). 
The remaining 10 outputs are listed by correlation strength: 

- Alcohol
- Volatile Acidity
- Sulphates
- Citric Acid
- Total Sulfur Dioxide
- Density
- Chlorides
- Fixed Acidity
- pH
- Free Sulfur Dioxide

Of all 10, alcohol had the strongest correlation of 0.48, while free sulfur 
dioxide had the weakest significant correlation of -0.051. 
Also, of note, is that there was a lack of wines in the range of 0.125 and 
0.25 g / dm^3 of citric acid only for quality levels 7 and 8. 
This may be caused by different categories of high-quality red wine purposefully 
have higher or lower levels of citric acid. 
As you may recall, citric acid can add ‘freshness’ and flavor to wine. Another 
interesting point is that most high-quality red wines (quality of 7 or 8) seems 
have sulphates levels between 0.65 and 0.82 g / dm^3.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There where 4 interesting relationships, 2 of which had weak negative 
correlations, volatile acidity and sulphates, volatile acidity and alcohol. 
Citric acid and sulphates had a weak positive correlation, while volatile 
acidity and citric acid had a moderate negative correlation.

### What was the strongest relationship you found?

The strongest relationship was a negative linear relationship between volatile 
acidity and citric acid, with a correlation of -0.55.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_set_up}
 
# some set up
wine$quality.cut <- cut(wine$quality, c(0,5,10)) 

```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# volatile.acidity and citric.acid
p1 <- ggplot(data = wine, 
             aes(x = volatile.acidity, y = citric.acid, 
                 color = quality.factor)) +
  geom_jitter(alpha = .75) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2)))

p2 <- ggplot(data = subset(wine, !is.na(quality.cut)), 
             aes(x = volatile.acidity, y = citric.acid, color = quality.cut)) +
  geom_smooth() +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(ylim = c(0,1))

grid.arrange(p1,p2, ncol = 2)

```
The plots are of volatile acidity, citric acid, and quality.

You can clearly see that higher quality wines tend to have low levels of 
volatile acidity, and slightly higher levels of citric acid, than lesser wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# volatile.acidty and sulphates
p1 <- ggplot(data = wine, 
             aes(x = volatile.acidity, y = sulphates, color = quality.factor)) +
  geom_jitter(alpha = .5) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(ylim = c(0.25,2, 0.5))

p2 <- ggplot(data = subset(wine, !is.na(quality.cut)), 
             aes(x = volatile.acidity, y = sulphates, color = quality.cut)) +
  geom_smooth() +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(ylim = c(0.25,2, 0.5))

grid.arrange(p1,p2, ncol = 2)

```

The plots are of volatile acidity, sulphates, and quality.

The trend of higher quality wines have less volatile acidity still holds. 
Furthermore, you can also see that sulphates tend to have higher levels in the 
better-quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# volatile.acidty and alcohol
p1 <- ggplot(data = wine, 
             aes(x = volatile.acidity, y = alcohol, color = quality.factor)) +
  geom_jitter(alpha = .75) +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2)))

p2 <- ggplot(data = subset(wine, !is.na(quality.cut)), 
             aes(x = volatile.acidity, y = alcohol, color = quality.cut)) +
  geom_jitter(alpha = 0) +
  geom_smooth() +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2)))

grid.arrange(p1,p2, ncol = 2)

```

The plots are of alcohol, volatile acidity, and quality.

The plot continues to show the trend of higher quality wines have less volatile 
acidity, but is also shows that those same wines tend to have higher alcoholic 
levels.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# citric.acid and sulphates
p1 <- ggplot(data = wine, 
             aes(x = citric.acid, y = sulphates, color = quality.factor)) +
  geom_jitter(alpha = .5) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2)))

p2 <- ggplot(data = subset(wine, !is.na(quality.cut)), 
             aes(x = citric.acid, y = sulphates, color = quality.cut)) +
  geom_jitter(alpha = 0) +
  geom_smooth() +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(reverse = T, 
                     override.aes = list(alpha = 1, size = 2)))

grid.arrange(p1,p2, ncol = 2)

```

These plots are of citric acid, sulphates, and quality.

The trend presented in this plot is that higher quality wines tend to have more 
sulphates present than lower quality wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Linear_model}

# Model, updated by variable
m1 <- lm(quality ~ alcohol, data = wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + sulphates)
m4 <- update(m3, ~ . + citric.acid)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
m6 <- update(m5, ~ . + density)
m7 <- update(m6, ~ . + chlorides)
m8 <- update(m7, ~ . + fixed.acidity)
m9 <- update(m8, ~ . + pH)
m10 <- update(m9, ~ . + free.sulfur.dioxide)

# display the linear model
mtable(m2,m4,m6,m8,m10)

```
Linear model utilizing 10 out of the 11 input variables.  
Model 10 (m10) is the final model with the highest R-squared and AIC value.

R-Squared tells us the proportion of variation in the dependent variable that 
has been explained by this model.
Typically, we want to see a R-squared of 0.7 or greater, but we don’t 
necessarily discard a model based on a low R-Squared value. 
Its a better practice to look at the AIC and prediction accuracy on validation 
sample when deciding on the efficacy of a model.
The Akaike’s information criterion - AIC (Akaike, 1974) measures the goodness 
of fit of an estimated statistical model and can also be used for model 
selection.
The lower the AIC value is the better.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

A few relationships observed were:

- Higher quality wines tended to have lower levels of volatile acidty.
- Higher quality wines tended to have slightly higher levels of citric acid 
present.
- Higher quality wines tended to have more alcoholic content. 
- Higher quality wines tended to have slightly higher levels of sulphates 
present.

When comparing volatile acidity, alcohol, and quality, the correlations between 
higher quality wines have higher alcoholic content and less volatile acidity 
became evident.

### Were there any interesting or surprising interactions between features?

One surprising feature was that higher quality tended to have slightly higher 
levels of sulphates, which is salt. This is surprising because it is breaking a 
preconception that salter wines may be of higher quality. That being said, by 
more salt, we are talking about a tenth (0.1) of a g / dm^3, which is a very 
small amount.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

We did create a linear model. The best fitted model created had a R-squared 
value of 0.360 and an AIC value of 3163. 
The largest limitation this model faces is that it quality was determined by 
taste testers.
While expert taste testers know a lot more about wine quality than I do, it 
still presents an element of human.
Furthermore, quality in of its self is may not directly relate to an enjoyable 
wine for your average consumer. 

This model is statistically significant with a P value less than 0.05. While 
we would prefer a higher R-squared value, model 10 (m10) has the highest value 
available.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

# Calculate correlation values
data <- cor(wine[,1:11], wine[,12], method = 'pearson')

# convert to dataframe
data <- as.data.frame((data))

# convert rownames to column
data <- tibble::rownames_to_column(data, 'name')

# plot
ggplot(data = data, aes(x = reorder(name, abs(V1)), y = V1, fill = abs(V1))) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  geom_label(aes(label = round(data$V1, 3))) +
  scale_fill_gradient(low = 'red', high = 'green') +
  guides(fill=FALSE) + # removes legend
  labs(x = 'Input Variables', y = 'Pearson Correlation',
       title = 'Correlation between Quality and Input Variables',
       subtitle = 'Input variables are based on physicochemical test from Red 
       Wine dataset.') +
  geom_vline(aes(xintercept = 1.5), colour= '#999999', linetype = 2) +
  geom_text(aes(x = 1.25, y = 0.2525, 
                label = 'Statistically Insignificant'), colour = '#999999')

```

### Description One

This plot is of the correlation values between quality and all of the input 
variables. 

It was chosen as one of the 3 final plots because it enables the reader to grasp 
the full range of the different correlation values between quality and the input 
variables.
It is important for the reader to understand the different correlation values.
A correlation is a statistical measurement that suggest the level of linear 
dependence between 2 variables.
In other words, we can use this as a ruff measurement of which variables we 
should pay the most attention to (such as alcohol) or which variables will 
probably be of little importance (such as residual sugar).
These values can range from -1 to +1, and the closer the value is to 0 the 
weaker the correlation is.
Furthermore, we are considering any value equal to and less than 0.05 to be 
statistically insignificant.


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

# plot
ggplot(data = wine, aes(x = quality, y = alcohol)) +
  geom_boxplot(aes(group=wine$quality.factor), color = '#000000') +
  geom_jitter(alpha = 1/3, color = 'orange') +
  scale_x_continuous(breaks = seq(3,8,1)) +
  geom_smooth(method = 'lm', alpha = 1/5, color = '#666666') +
  labs(x = 'Quality', y = 'Alcohol (% by volume)', 
       title = 'Alcoholic Content by Quality Level')

```

### Description Two

This is a jitter plot of alcohol content by quality level with a boxplot and 
linear regression line overlay.

It was chosen because it clearly presents the relationship between alcohol and 
quality, the strongest correlation of all the input variables.
The jitter plot is a scatter plot that adds a small amount of randomness to the 
discrete position (quality), which helps to avoid overpotting, and more clearly
show concentrations.
The box plot is a standardized way of displaying the distribution of the data, 
and the linear regression line allows us to see the trend that the correlation 
value was calculating the level of linear dependence of. 

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height= 5, fig.width= 11}

# jitter plot
p1 <- ggplot(data = wine, 
             aes(x = volatile.acidity, y = alcohol, color = quality.factor)) +
  geom_jitter(alpha = .75) +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality', reverse = T, 
                     override.aes = list(alpha = 1, size = 2))) +
  labs(x = 'Volatile Acidity (acetic acid - g / dm^3)', 
       y = 'Alcohol (% by volume)',
       title = 'Alcohol Content by Volatile Acidity', 
       subtitle = 'With Quality Scores')

# smooth line plot
p2 <- ggplot(data = subset(wine, !is.na(quality.cut)), 
             aes(x = volatile.acidity, y = alcohol, color = quality.cut)) +
  geom_jitter(alpha = 0) +
  geom_smooth(method = 'loess') +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Quality', reverse = T, 
                     override.aes = list(alpha = 1, size = 2))) +
  labs(x = 'Volatile Acidity (acetic acid - g / dm^3)', 
       y = 'Alcohol (% by volume)',
       title = 'Alcohol Content by Volatile Acidity', 
       subtitle = 'With Quality Score Lines')

# grid arrange
grid.arrange(p1,p2, ncol = 2)

```

### Description Three

This is a jitter and smooth line plot of volatile acidity, alcohol, and quality.

It was chosen because it allows us to see the multivariable trends between 
volatile acidity, alcohol, and quality.
The jitter plot allows us to see the concentrations and quality level, thus 
enabling use to visually determine if any trends are present. 
The smooth line plot is another tool that is very effective in allowing us to 
see what the trend actually looks like.
However, due to its nature, it can over generalize, although that can be 
mitigated through the use of the confidence bands.

------

# Reflection

The Red Wine Quality dataset was a tidy dataset with almost 1,600 wines, 11 
physicochemical (input) variables, and 1 sensory (output) variable. 
The majority of the wines where of a normal quality with very few being 
considered of high or low quality. 
This is the greatest limitation in the dataset because it made it much more 
difficult to identify chemical trends that could be used to predict high- or 
low-quality wines. 
The majority of the wines where not considered sweet, where considered acidic 
(pH score between 3.2 and 3.4) and had an alcohol content between 9.5 and 11.1%.
The majority of the input variables had weak correlation quality, with the 
strongest correlation being alcohol, followed by volatile acidity.
Residual sugar was the only variable to have a statically insignificant 
correlation to quality.
It was found that higher quality wines where more likely to have higher alcohol 
content, and lower volatile acidity levels.
Furthermore, higher quality wines also tended to have slightly higher citric 
acid and sulphates levels.
Lastly, when comparing alcohol content, volatile acidity, and quality, you can 
clear see the concentration of higher quality wines.
All of these findings indicate that alcohol and volatile acidity would be the 
best 2 indicators of wine quality, a finding supported by the linear regression 
model which gave these 2 variables high significant ratings. 

When starting this analysis, my first thought would be that this would be easy. 
Primarily, I would only need to consider correlations to quality, make some 
nice plots, and submit. 
What I was not expecting was the vast number of low correlations to quality, 
and intercorrelation between input variables, such as free and total sulfur 
dioxide. 
This called for a more in-depth exploration of the data. 
The box plots and linear regression line overlays worked well on viewing this 
behavior. Furthermore, the smooth lines in the multivariable plots help to spot 
tends. 
However, that brings me the last issue experienced, that is the fact that 
discrete variables can be rather difficult to analyze. 
If you put a discrete variable (like quality) into an ordered pair with 2 
levels, then can begin to overgeneralize. 
However, if you do not, then it may be rather had to identify any trends.

As already mentioned the greatest limitation was the lack of high- and 
low-quality wines in the dataset.
However, another limitation is that the quality was determined by human sensory 
data. 
While measures were taken to limit the human error, averaging the quality 
rating from 3 separate testers, it will always remain present in the dataset. 
The best way to improve this analysis is to obtain more data entrees of high- 
and low-quality wines.
This would balance out the quality variables, allowing for more accurate trends 
to be identified, regarding high- and low-quality wines.
